---
title: "Pediatric ALL microbiota: quantitative and structured synthesis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  input_xlsx: "ALL_microbiota_ultimate_input.xlsx"
  output_dir: "analysis_outputs"
---

```{r setup, include=FALSE}
required_packages <- c(
  "tibble",
  "readxl", "dplyr", "tidyr", "ggplot2", "stringr",
  "metafor", "openxlsx", "knitr"
)
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    paste0(
      "Install these packages before rendering: ",
      paste(missing_packages, collapse = ", "),
      "\nRun while online: install.packages(c(",
      paste(sprintf('"%s"', missing_packages), collapse = ", "),
      "))"
    )
  )
}

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(metafor)
library(openxlsx)
library(knitr)

input_xlsx <- params$input_xlsx
output_dir <- params$output_dir
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

save_both <- function(plot, stem, width = 11, height = 7) {
  ggsave(file.path(output_dir, paste0(stem, ".png")), plot,
         width = width, height = height, dpi = 300, bg = "white")
  ggsave(file.path(output_dir, paste0(stem, ".pdf")), plot,
         width = width, height = height, device = "pdf", bg = "white")
}

read_sheet <- function(sheet) read_excel(input_xlsx, sheet = sheet)

study_master <- read_sheet("Study_master")
risk_of_bias <- read_sheet("Risk_of_bias")
cohort_overlap <- read_sheet("Cohort_overlap")
findings_all <- read_sheet("Findings_all")
heatmap_input <- read_sheet("Heatmap_input")
taxon_direction <- read_sheet("Taxon_direction")
ma_shannon <- read_sheet("MA_Shannon")
ma_richness <- read_sheet("MA_Richness")

# Fixed interpretation used throughout.
direction_levels <- c("Lower/depleted", "No difference", "Higher/enriched", "Different/mixed", "Not assessed")
direction_colours <- c(
  "Lower/depleted" = "#3B82F6",
  "No difference" = "#D1D5DB",
  "Higher/enriched" = "#EF4444",
  "Different/mixed" = "#F59E0B",
  "Not assessed" = "#FFFFFF"
)
```


> **Repository version.** This file reproduces the exploratory alpha-diversity
> meta-analyses and descriptive structured-synthesis figures used for the manuscript.
> Obsolete inferential vote-counting/sign-test outputs and aggregate risk-of-bias
> scores are intentionally not generated.

# 1. Validation

```{r validation}
validation <- list()
validation[["duplicate_finding_ids"]] <- findings_all %>% count(finding_id) %>% filter(n > 1)
validation[["invalid_direction_codes"]] <- bind_rows(
  heatmap_input %>% transmute(source = "Heatmap_input", id = paste(row_label, feature), direction_code),
  taxon_direction %>% transmute(source = "Taxon_direction", id = taxon_id, direction_code)
) %>% filter(!direction_code %in% c(-1, 0, 1, 2, 9))
validation[["main_shannon_incomplete"]] <- ma_shannon %>%
  filter(include_main) %>%
  filter(if_any(c(n_all, mean_all, sd_all, n_control, mean_control, sd_control), is.na))
validation[["main_richness_incomplete"]] <- ma_richness %>%
  filter(include_main) %>%
  filter(if_any(c(n_all, mean_all, sd_all, n_control, mean_control, sd_control), is.na))
validation[["cohort_overlap"]] <- cohort_overlap

validation_summary <- tibble(
  check = names(validation),
  n_rows = vapply(validation, nrow, integer(1))
)
kable(validation_summary)
```

# 2. Exploratory alpha-diversity meta-analyses

The main analyses include three diagnosis-time studies. Negative Hedges' g indicates lower diversity or richness in ALL. The extended analyses add De Pietri as an early-induction sensitivity analysis.

```{r meta-functions}
run_smd_meta <- function(data, include_column, analysis_name, file_stem) {
  dat <- data %>% filter(.data[[include_column]] %in% TRUE)
  if (nrow(dat) < 2) stop(paste("Fewer than two studies for", analysis_name))

  es <- metafor::escalc(
    measure = "SMD",
    m1i = mean_all, sd1i = sd_all, n1i = n_all,
    m2i = mean_control, sd2i = sd_control, n2i = n_control,
    data = dat, vtype = "UB"
  ) %>%
    mutate(
      ci_low = yi - qnorm(0.975) * sqrt(vi),
      ci_high = yi + qnorm(0.975) * sqrt(vi),
      analysis = analysis_name
    )

  model <- metafor::rma(
    yi = yi, vi = vi, data = es,
    method = "REML", test = "knha",
    slab = study_id
  )
  prediction <- predict(model)

  summary <- tibble(
    analysis = analysis_name,
    k = model$k,
    pooled_hedges_g = as.numeric(model$b),
    ci_low = model$ci.lb,
    ci_high = model$ci.ub,
    p_value = model$pval,
    tau2 = model$tau2,
    I2 = model$I2,
    Q = model$QE,
    Q_p = model$QEp,
    prediction_low = prediction$pi.lb,
    prediction_high = prediction$pi.ub,
    estimator = "REML",
    inference = "Hartung-Knapp"
  )

  draw_forest <- function(device_file, device = c("png", "pdf")) {
    device <- match.arg(device)
    if (device == "png") {
      png(device_file, width = 2500, height = 1800, res = 300)
    } else {
      pdf(device_file, width = 8.5, height = 6.5, useDingbats = FALSE)
    }
    par(mar = c(4, 4, 2, 2))
    metafor::forest(
      model,
      xlab = "Hedges' g (negative = lower in ALL)",
      refline = 0,
      header = c("Study", "Hedges' g [95% CI]"),
      cex = 0.9
    )
    title(analysis_name)
    dev.off()
  }

  draw_forest(file.path(output_dir, paste0(file_stem, ".png")), "png")
  draw_forest(file.path(output_dir, paste0(file_stem, ".pdf")), "pdf")

  loo <- if (model$k >= 3) as.data.frame(metafor::leave1out(model)) %>%
    tibble::rownames_to_column("omitted_study") else tibble()

  list(data = es, model = model, summary = summary, leave_one_out = loo)
}

shannon_main <- run_smd_meta(ma_shannon, "include_main", "Shannon diversity at diagnosis", "forest_Shannon_main")
shannon_extended <- run_smd_meta(ma_shannon, "include_extended", "Shannon diversity: extended early-induction sensitivity", "forest_Shannon_extended")
richness_main <- run_smd_meta(ma_richness, "include_main", "Chao1 richness at diagnosis", "forest_Richness_main")
richness_extended <- run_smd_meta(ma_richness, "include_extended", "Richness: extended sensitivity including observed species", "forest_Richness_extended")

meta_summary <- bind_rows(
  shannon_main$summary, shannon_extended$summary,
  richness_main$summary, richness_extended$summary
)
kable(meta_summary, digits = 3)
```

# 3. Direction heatmaps

```{r direction-heatmaps}
heatmap_plot <- function(panel_name) {
  d <- heatmap_input %>%
    filter(panel == panel_name) %>%
    mutate(
      direction_label = factor(direction_label, levels = direction_levels),
      row_label = factor(row_label, levels = rev(unique(row_label))),
      feature = factor(feature, levels = unique(feature))
    )

  p <- ggplot(d, aes(x = feature, y = row_label, fill = direction_label)) +
    geom_tile(color = "white", linewidth = 0.7) +
    geom_text(aes(label = cell_label), size = 3.2) +
    scale_fill_manual(values = direction_colours, drop = FALSE) +
    labs(title = paste("Direction heatmap:", panel_name), x = NULL, y = NULL, fill = "Direction") +
    theme_minimal(base_size = 11) +
    theme(
      legend.position = "bottom",
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 40, hjust = 1),
      axis.text.y = element_text(face = "bold")
    )
  p
}

heatmap_panels <- unique(heatmap_input$panel)
heatmap_plots <- lapply(heatmap_panels, heatmap_plot)
names(heatmap_plots) <- heatmap_panels
for (nm in names(heatmap_plots)) {
  stem <- paste0("heatmap_", str_replace_all(str_to_lower(nm), "[^a-z0-9]+", "_"))
  save_both(heatmap_plots[[nm]], stem, width = 12, height = 6.5)
}
```

# 4. Taxon-direction heatmap

```{r taxon-heatmap}
taxon_plot_data <- taxon_direction %>%
  mutate(
    direction_label = factor(direction_label, levels = direction_levels),
    study_context = paste(study_id, clinical_context, sep = " - "),
    study_context = factor(study_context, levels = rev(unique(study_context))),
    canonical_taxon_or_group = factor(canonical_taxon_or_group, levels = unique(canonical_taxon_or_group))
  )

p_taxa <- ggplot(taxon_plot_data,
                 aes(x = canonical_taxon_or_group, y = study_context, fill = direction_label)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(direction_code == -1, "↓", ifelse(direction_code == 1, "↑", "≠"))), size = 3) +
  facet_wrap(~ panel, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = direction_colours, drop = FALSE) +
  labs(title = "Direction of selected recurrent or clinically interpretable taxa",
       subtitle = "Only transparent, curated taxa are shown; granular results remain in Findings_all",
       x = NULL, y = NULL, fill = "Direction") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 7),
    strip.text = element_text(face = "bold")
  )
save_both(p_taxa, "taxon_direction_heatmap", width = 14, height = 11)
```


# 5. Export numerical results and audit tables

This export contains the numerical meta-analysis results used in the manuscript,
the study-level effect-size data, leave-one-out sensitivity analyses, cohort-overlap
audit, curated descriptive-synthesis inputs, and the raw risk-of-bias table.

No vote-counting p-values, weighted sign tests, pooled directional scores, funnel plots,
publication-bias tests, or meta-regression are calculated.

```{r export-results}
results <- list(
  Meta_summary = meta_summary,
  Shannon_main_studies = shannon_main$data,
  Shannon_extended_studies = shannon_extended$data,
  Richness_main_studies = richness_main$data,
  Richness_extended_studies = richness_extended$data,
  Shannon_main_leave1out = shannon_main$leave_one_out,
  Shannon_ext_leave1out = shannon_extended$leave_one_out,
  Richness_main_leave1out = richness_main$leave_one_out,
  Richness_ext_leave1out = richness_extended$leave_one_out,
  Study_master = study_master,
  Cohort_overlap = cohort_overlap,
  Risk_of_bias = risk_of_bias,
  Heatmap_input = heatmap_input,
  Taxon_direction = taxon_direction,
  Validation_summary = validation_summary
)

openxlsx::write.xlsx(
  results,
  file.path(output_dir, "analysis_results.xlsx"),
  overwrite = TRUE
)

manifest <- tibble(
  file = list.files(output_dir, full.names = FALSE),
  generated = Sys.time()
)

write.csv(
  manifest,
  file.path(output_dir, "analysis_manifest.csv"),
  row.names = FALSE
)

writeLines(
  capture.output(sessionInfo()),
  file.path(output_dir, "sessionInfo.txt")
)
```


Render with:

```r
rmarkdown::render("ALL_microbiota_analysis_GitHub_final.Rmd")
```

Outputs are written to `analysis_outputs/`.
